---
title: "Data Analytics Basic Methods"
output: pdf_document
---
# Assignment 1 (10%)

[Zion Yitbarek]

[Insert course section & student number]

This assignment can be submitted using either Python or R, whichever you prefer.
If using R, you must submit an RMD file with its knitted file (PDF or HTML). To learn more about
knitting and R markdown, visit R Markdown.

If using Python, you must submit an IPYNB file and its exported PDF/HTML with clearly printed/shown
answers.

Failing to submit both files ({RMD + knitted PDF/HTML} OR {IPYNB + PDF/HTML}) will be subject to a 30%
mark deduction.

NOTE: IF YOU USE R STUDIO , YOU SHOULD NEVER HAVE install.packages IN YOUR CODE;
OTHERWISE, THE Knit OPTION WILL RAISE AN ERROR. COMMENT OUT ALL PACKAGE INSTALLATIONS
BUT KEEP library() CALLS.

NOTE: If you answer the questions in R, all your answers should be in R (ignore Python questions). If you
answer the questions in Python, all your answers


## Question 1 (40 Points)

### Q. 1a (5 points)
Create a vector of all even numbers from 3 to 49.
Hint: Use seq() in R or range() in python

```{r}
seq(4,48,2)
```

### Q. 1b (5 points)
Create and print a vector `x` with all integers from 1 to 80, and a vector `y` with odd integer numbers in the same range.
Hint: use `seq()`function in R or range()in python. Calculate the difference in lengths of the vectors `x` and `y`. Hint: use length() in R or len() in python

```{r}
x <-seq(1,80,1)
y <-seq(1,79,2)
print(x)
print (y)
```

### Q. 1c (10 points)
Create a new vector, "x_square", with the square of elements at indices 3 5 7 9 11 13 15 17 19 21 23 25 27 29 31 33 35 37 39 41 from the variable "x". (Hint: Use indexing rather than a for loop.) 

Calculate the mean and median of the LAST 7 values from x_square.

```{r}
indices <- c(3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35 ,37,39,41)
x_square <- x[indices]^2
print(x_square)
x_square_new <- tail(x_square,7)

mean_x_square <-mean(x_square_new)
median_x_square <- median(x_square_new)

print(paste("The mean is" , mean_x_square))
print(paste("The median is" , median_x_square))
```


### Q. 1d (10 Points)
For a given factor variable
factorVar <- factor(c("3.5", "8.1", "5.2", "9.8")) in R or
factorVar = ["3.5", "8.1", "5.2", "9.8"] in python
Please convert it into numeric data type and show proof that it is converted to numeric.
```{r}
factorVar <- factor(c(3.5, 8.1, 5.2, 9.8))
numvar <- as.numeric(as.character(factorVar))
str(numvar)
print(numvar)
```

### Q. 1e (10 points)
A comma-separated values file A comma-separated values file `dataset.csv` consists of missing values represented by Not a Number (`NaN`) and question mark (`?`). How can you read this type of files in R? consists of missing values represented by Not a Number (`NaN`) and question mark (`?`). How can you read this type of files in R?

```{r}
getwd() 
file_1<-"C:/Users/user/Documents"
data_1 <- read.csv(file.choose(), na.strings = c("NaN", "?"))
print(data_1)

```

## Question 2 (60 Points)
Compute with using R or Python commands

### Q. 2a (10 points)
$$\sum_{n=1}^{100}\frac{4^{n}}{(n+1)!}$$ 
Hint: Use `factorial(n)` to compute $n!$ 

```{r}
numseq <- seq(1, 100, 1)
numseq2 <- 4^numseq
fact_seq <- factorial(numseq +1)
subtotal <- numseq2/ fact_seq
final_answer <- sum(subtotal)
print(final_answer)

```

### Q. 2b (10 points)
 $$\sum_{n=1}^{25}\left(\frac{3^{n}}{n^3} + \frac{n^{4}}{4^{n}}\right)$$
```{r}
seq_num_1 <- seq(1, 25 ,1)
seq_num_2 <- 3^seq_num_1
seq_num_3 <- seq_num_1^3
term1 <-seq_num_2/seq_num_3
term2 <-seq_num_1^4/4^seq_num_1
final <- sum(term1 +term2)
print(final)

```
### Q. 2c (10 points)
$$\sum_{n=0}^{100} \frac{(n+1)5^n}{25^{n+1}}$$

```{r}
seq_num_a <-seq(0,100,1)
numerator <- (seq_num_a +1) *5^seq_num_a
denominator <- 25^(seq_num_a +1)
subans <- numerator/denominator
finalans <- sum(subans)
print(finalans)


```

### Q. 2d (10 points)

$$\prod_{n=2}^{24} \left(3n + \frac{3}{\sqrt[4]{n}}\right)$$

```{r}
seq_num_b <- seq(2,24,1)

num1= 3*seq_num_b +3
num2 <- 3/(seq_num_b^(1/4))
num3 <- num1 +num2

finalnum <-prod(num3)
print(finalnum)
```

### Q. 2e (10 points)

$$\sum_{n=1}^{9}\left(\frac{3^{n}}{n^3} + \frac{n^{5}}{4^{5}}+\frac{n^{7}}{7^{n}}+\frac{n^{9}}{9^{n}}\right)$$(5 points)

```{r}
final_seq <- seq(1,9,1)

part1 <-((3^final_seq)/(final_seq^3))
part2 <-part2 <- ((final_seq^5)/(4^5))
part3 <- ((final_seq^7)/(7^final_seq))
part4 <- ((final_seq^9)/(9^final_seq))

final_sum <- part1 + part2+ part3+ part4

final_total <- sum(final_sum)
print(final_total)
```

### Q. 2f (10 points)

Describe the purpose of is.logical() , is.character() , is.numeric() , and is.na() functions in R.

In Python, describe the purpose of isinstance() , type() , and pd.isna() functions from the pandas library.

Hint: Please use x <- c("a", FALSE, "b", NA, 2, TRUE) in R or x = ["a", False, "b", None, 2, True] in Python to explain your description.

```{r}
#Creating the vector
x<-c("a", FALSE, "b", NA, 2, TRUE)

#The purpose of is.logical() is to check if the value/s is logical, meaning that if the value/s are true or false.
is_logical <- is.logical(x)
print(is_logical)
```


```{r}
#The purpose of  is.character() is to check if the values/s is a string
is_character <- is.character(x)
print(is_character)
#Final result will be TRUE for "a" and "b" and FALSE for everything else
```


```{r}
# The purpose of is.numeric(): Checks if the value is numeric (number).
is_numeric <- is.numeric(x)
print(is_numeric)
#Final result will be TRUE for the fifth element (2) and FALSE for the others as they aren't numbers.
```


```{r}
#The purpose of is.na() is to check if there are any missing values (NA)
is_na <- is.na(x)
print(is_na)
#Final result is will be FALSE for the first 3 elements ("a", FALSE, "b"), TRUE for the fourth element and FALSE for the last two elements(2,TRUE)



```










